---
title: "Examine metadata"
author: "Kimberly Olney, Ph.D"
date: "10/21/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

This script will examine the variables in the metadata to determine if there are differences by disease type and/or sex differences for clinical data such as brain weight. 
# Setup
```{r setup, message=FALSE, warning=FALSE, tidy=TRUE}
knitr::opts_knit$set(root.dir = ".")
```

# User defined variables
sourcing file_paths_and_colours will load the libraries and user defined variables for colors of the graphs 
```{r set_variables, message=FALSE, warning=FALSE, tidy=TRUE}
source(here::here("scripts", "file_paths_and_colours.R"))
```

# Total percent male
```{r percent_male}
total_sex_percent <- metadata %>% 
  group_by(Sex) %>%
  dplyr::count() %>% 
  ungroup(Sex) %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
```

# APOE E4 allele count
```{r APOE_E4_allele_count, message=FALSE, warning=FALSE, tidy=TRUE}
metadata$APOE_E4_allele_count <-
  ifelse(
    metadata$APOE == "E2E3",
    0,
    ifelse(
      metadata$APOE == "E2E4",
      1,
      ifelse(
        metadata$APOE == "E3E3",
        0,
        ifelse(
          metadata$APOE == "E3E4",
          1,
          ifelse(
            metadata$APOE == "E4E4", 2, "NA")))))
metadata$APOE_E4_allele_count <- as.numeric(metadata$APOE_E4_allele_count)
metadata$sex_numeric <- ifelse(metadata$Sex == "male", 0, 1)
metadata$Race_numeric <- ifelse(metadata$Race == "Caucasian", 0, 1)
```

# Output updated metadata table
```{r metadata_output, message=FALSE, warning=FALSE, tidy=TRUE}
write.table(
  metadata,
  paste0(pathToRawData, "ATS_proteomics_metadata.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
# clean up
remove(APOE, metadata_ATS, df, RNA_metrics_metadata)
```

## Proteomics sample count
```{r bar_plot, message=FALSE, warning=FALSE, tidy=TRUE}
# create bar plot showing number of samples for each group
bar <- ggplot(metadata, aes(TYPE, after_stat(count), fill = TYPE)) +
  geom_bar() +
  theme_bw() +
  xlab("Disease group") +
  ggtitle("Proteomics sample count by disease group") +
  geom_text(
    stat = 'count',
    aes(label = after_stat(count)),
    vjust = 1.6,
    color = "white",
    size = 3.5
  ) +
    theme_bw() + 
  scale_fill_manual(values = TypeColors)
bar

# pie chart
data <- metadata %>%
  group_by(TYPE) %>%
  dplyr::count() %>%
  ungroup() %>%
  mutate(perc = `n` / sum(`n`)) %>%
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

pie <- ggplot(data, aes(x = "", y = perc, fill = TYPE)) +
  geom_col(color = "black") +
  geom_label(
    aes(label = labels),
    color = c("white"),
    position = position_stack(vjust = 0.5),
    show.legend = FALSE
  ) +
  guides(fill = guide_legend(title = "Disease group")) +
  scale_fill_manual(values = TypeColors) +
  coord_polar(theta = "y") +
  theme_void() +
  theme(legend.position = "none")
pie
```
save plot
```{r message=FALSE, warning=FALSE, tidy=TRUE}
ggarrange(bar,
          pie,
          ncol = 2,
          widths = c(2.5, 1))

path <- paste0("../results/metadata/TYPE.count")
saveToPDF(paste0(path, ".pdf"), width = 7.25, height = 3.5)

# clean up
remove(bar, pie, data)
```

## Split by sex 
```{r message=FALSE, warning=FALSE, tidy=TRUE}
bar_sex <- ggplot(metadata, aes(x = TYPE, fill = Sex)) + 
  geom_bar(aes(y = after_stat(count)), position = "dodge") + 
  theme_bw() +
  xlab("Disease group") + 
  ggtitle("Proteomics sample count by inferred sex") +
  geom_text(stat='count', aes(x = TYPE, label=after_stat(count)), 
            position = position_dodge(width = 1), 
              vjust=-.25, 
            color="black", size=3.5) + 
  scale_fill_manual(values = SexColors) +
  guides(fill = guide_legend(title = "Sex"))+
  scale_y_continuous(breaks = seq(0, 350, by = 100), limits = c(0, 350))
bar_sex

data <- metadata %>% 
  group_by(Sex,TYPE) %>%
  dplyr::count() %>% 
  ungroup(Sex) %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
```
```{r message=FALSE, warning=FALSE, tidy=TRUE}
bar_sex
path <- paste0("../results/metadata/TYPE.sex.count")
saveToPDF(paste0(path, ".pdf"), width = 6.88, height = 4.33)

# clean up
remove(bar_sex, data)
```

# A-T-S sample count
```{r message=FALSE, warning=FALSE, tidy=TRUE}
# create bar plot showing number of samples for each A-T-S group
bar <- ggplot(metadata, aes(group.redefined, after_stat(count), fill = group.redefined)) +
  geom_bar() +
  theme_bw() +
  xlab("Disease group") +
  ggtitle("Proteomics sample count by A-T-S scores") +
  theme_bw() + 
  geom_text(
    stat = 'count',
    aes(label = after_stat(count)),
    vjust = 1.6,
    color = "black",
    size = 3.5
  ) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 10, angle = 75, vjust = 1, hjust=1),
        axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.text=element_text(size=10),
        legend.title=element_text(size=10)) +
  scale_fill_manual(values = ATSColors)
bar
path <- paste0("../results/metadata/ATS.count")
saveToPDF(paste0(path, ".pdf"), width = 7, height = 5)
```

```{r message=FALSE, warning=FALSE, tidy=TRUE}
# pie chart
data <- metadata %>%
  group_by(group.redefined) %>%
  dplyr::count() %>%
  ungroup() %>%
  mutate(perc = `n` / sum(`n`)) %>%
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

pie <- ggplot(data, aes(x = "", y = perc, fill = group.redefined)) +
  geom_col(color = "black") +
  geom_label(
    aes(label = labels),
    color = c("black"),
    position = position_stack(vjust = 0.5),
    show.legend = FALSE
  ) +
  guides(fill = guide_legend(title = "Disease group")) +
  scale_fill_manual(values = ATSColors) +
  coord_polar(theta = "y") +
  theme_void() +
  theme(legend.position = "none")
pie
```

save plot
```{r message=FALSE, warning=FALSE, tidy=TRUE}
ggarrange(bar,
          pie,
          ncol = 2,
          widths = c(2.5, 1))

path <- paste0("../results/metadata/ATS.count.pie")
saveToPDF(paste0(path, ".pdf"), width = 9, height = 6)

# clean up
remove(bar, data, pie) # do not remove df, used for split by sex 
```

## A-T-S split by sex 
```{r message=FALSE, warning=FALSE, tidy=TRUE}
bar_sex <- ggplot(metadata, aes(x = group.redefined, fill = Sex)) + 
  geom_bar(aes(y = after_stat(count)), position = "dodge") + 
  theme_bw() + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.text=element_text(size=10),
        legend.title=element_text(size=10)) +
  xlab("Disease group") + 
  ggtitle("Proteomics sample count by inferred sex") +
  geom_text(stat='count', aes(x = group.redefined, label=after_stat(count)), 
            position = position_dodge(width = 1), 
              vjust=-.25, 
            color="black", size=3.5) + 
  scale_fill_manual(values = SexColors) +
  guides(fill = guide_legend(title = "Sex inferred"))+
  scale_y_continuous(breaks = seq(0, 125, by = 25), limits = c(0, 125))
bar_sex
```

```{r message=FALSE, warning=FALSE, tidy=TRUE}
bar_sex
path <- paste0("../results/metadata/ATS.sex.count")
saveToPDF(paste0(path, ".pdf"), width = 9, height = 7)

remove(bar_sex, pie_sex, data, df)
```
# Save metadata
```{r message=FALSE, warning=FALSE, tidy=TRUE}
saveRDS(metadata, paste0("../rObjects/ATS_metadata.rds"))
```

# Categorial variables
### TYPE
```{r message=FALSE, warning=FALSE, tidy=TRUE, eval=FALSE}
metadata_categorical <-
  data.frame(
    metadata$LBD.type,
    metadata$AD.subtype,
    metadata$CDLB,
    metadata$TDP.type,
    metadata$APOE,
    metadata$MAPT,
    metadata$GRN,
    metadata$TMEM106b,
    metadata$Braak.NFT,
    metadata$Thal.amyloid,
    metadata$MF.Tau,
    metadata$MF.Amyloid,
    metadata$VaD,
    metadata$TDP.43,
    metadata$FHx,
    metadata$MF.SP,
    metadata$MF.NFT,
    metadata$MF.LB,
    metadata$Cing.LB,
    metadata$Cing.Synuclein,
    metadata$Race
  )
# To do it for all names
column_variables <-
  c(
    "LBD.type",
    "AD.subtype",
    "CDLB",
    "TDP.type",
    "APOE",
    "MAPT",
    "GRN",
    "TMEM106b",
    "Braak.NFT",
    "Thal.amyloid",
    "MF.Tau",
    "MF.Amyloid",
    "VaD",
    "TDP.43",
    "FHx",
    "MF.SP",
    "MF.NFT",
    "MF.LB",
    "Cing.LB",
    "Cing.Synuclein",
    "Race"
  )

bar_plot_fun <- function(i, j) {
  ggplot(metadata, aes(TYPE, after_stat(count), fill = factor(i))) +
    geom_bar(colour = "black") +
    theme_bw() +
    ggtitle(j) + 
    guides(fill = guide_legend(title = j)) +
    xlab("Disease group") +
    ylab("Count") +
    scale_fill_manual(values=colorbindColors) +
    geom_text(
      stat = 'count',
      aes(label = after_stat(count)),
      position = position_stack(vjust = .5),
      color = "black",
      size = 3.5
    )

}
bar_plots <-
  Map(bar_plot_fun, i = metadata_categorical, j = column_variables)
bar_plots


p <- list()
for (i in 1:length(column_variables)) {
  p[[i]] <- grid.arrange(grobs = c(bar_plots[i]))
}

plotnames = imap(bar_plots, ~ paste0("../results/metadata/TYPE.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames,
      p,
      ~ ggsave(
        filename = .x,
        plot = .y,
        height = 5,
        width = 6.5
      ))
```

```{r message=FALSE, warning=FALSE, tidy=TRUE}
# clean up
remove(bar_plots, pie_plots, plotnames, p) # do not remove metadata_categorical
```


### A-T-S
```{r message=FALSE, warning=FALSE, tidy=TRUE}
bar_plot_fun <- function(i, j) {
  ggplot(metadata, aes(group.redefined, after_stat(count), fill = factor(i))) +
    geom_bar(colour = "black") +
    theme_bw() +
    ggtitle(j) + 
    guides(fill = guide_legend(title = j)) +
    xlab("Disease group") +
    ylab("Count") +
    scale_fill_manual(values=colorbindColors) +
    geom_text(
      stat = 'count',
      aes(label = after_stat(count)),
      position = position_stack(vjust = .5),
      color = "black",
      size = 3.5
    ) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
}
bar_plots <-
  Map(bar_plot_fun, i = metadata_categorical, j = column_variables)
bar_plots

p <- list()
for (i in 1:length(column_variables)) {
  p[[i]] <- grid.arrange(grobs = c(bar_plots[i]), heights = c(2,1))
}

plotnames = imap(bar_plots, ~ paste0("../results/metadata/ATS.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames,
      p,
      ~ ggsave(
        filename = .x,
        plot = .y,
        height = 8.5,
        width = 9.5
      ))
```

```{r message=FALSE, warning=FALSE, tidy=TRUE}
# clean up
remove(bar_plots, plotnames, p) # do not remove metadata_categorical
```

# Continuous variables 
### TYPE
```{r message=FALSE, warning=FALSE, tidy=TRUE}
metadata_continuous <-
  data.frame(
    metadata$Brain.wt,
    metadata$Duration,
    metadata$Age,
    metadata$PMI
  )
column_variables <-
  c(
    "Brain.wt",
    "Duration",
    "Age",
    "PMI"
  )
TYPE_comparison <-
  list(
    c("CONTROL", "LBD")
  )

violin_plot_fun <- function(i, j) {
  ggplot(metadata, aes(TYPE, i, fill = TYPE)) +
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.shape = NA) + 
    geom_jitter(shape=16, position=position_jitter(0.2)) +
    theme_bw() + 
    ggtitle(j) +
    xlab("Disease group") +
    ylab(j) +
    stat_compare_means(comparisons = TYPE_comparison) +
    scale_fill_manual(values=TypeColors) +
    theme(legend.position = "none")
}
violin_plots <-
  Map(violin_plot_fun, i = metadata_continuous, j = column_variables)
violin_plots

plotnames = imap(violin_plots, ~paste0("../results/metadata/TYPE.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames, violin_plots, ~ggsave(filename = .x, plot = .y, 
                                    height = 4, width = 5.5))

remove(violin_plots) 
```

## TYPE.sex
```{r message=FALSE, warning=FALSE, tidy=TRUE}
violin_plot_fun <- function(i, j) {
  ggplot(metadata, aes(TYPE, i, fill = TYPE)) +
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.shape = NA) + 
    geom_jitter(shape=16, position=position_jitter(0.2)) +
    theme_bw() + 
    ggtitle(j) +
    xlab("Disease group") +
    ylab(j) +
    stat_compare_means(comparisons = TYPE_comparison) +
    scale_fill_manual(values=TypeColors) +
    theme(legend.position = "none") +
    facet_grid(. ~ Sex)
}
violin_plots <-
  Map(violin_plot_fun, i = metadata_continuous, j = column_variables)
violin_plots

plotnames = imap(violin_plots, ~paste0("../results/metadata/TYPE.Sex.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames, violin_plots, ~ggsave(filename = .x, plot = .y, 
                                    height = 4, width = 7.5))
remove(violin_plots) 
```


### A-T-S
```{r message=FALSE, warning=FALSE, tidy=TRUE}
metadata$group.redefined
ATS_comparison <-
  list(
    c("CONTROL", "LBD_ATS"),
    c("CONTROL", "LBD_AS"),
    c("CONTROL", "LBD_TS"),
    c("CONTROL", "LBD_S")

  )
violin_plot_fun <- function(i, j) {
  ggplot(metadata, aes(group.redefined, i, fill = group.redefined)) +
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.shape = NA) + 
    geom_jitter(shape=16, position=position_jitter(0.2)) +
    theme_bw() + 
    ggtitle(j) +
    xlab("Disease group") +
    ylab(j) +
    stat_compare_means(comparisons = ATS_comparison) +
    scale_fill_manual(values=ATSColors) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
}
violin_plots <-
  Map(violin_plot_fun, i = metadata_continuous, j = column_variables)
violin_plots

plotnames = imap(violin_plots, ~paste0("../results/metadata/ATS.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames, violin_plots, ~ggsave(filename = .x, plot = .y, 
                                    height = 7, width = 7))
remove(violin_plots) 
```


### APOE
```{r message=FALSE, warning=FALSE, tidy=TRUE}
metadata$APOE <- as.factor(metadata$APOE)
metadata_continuous <-
  data.frame(
    metadata$Brain.wt,
    metadata$Duration,
    metadata$Age,
    metadata$Cing.LB,
    metadata$MF.LB,
    metadata$Braak.NFT,
    metadata$Thal.amyloid,
    metadata$MF.SP,
    metadata$MF.NFT,
    metadata$MF.Amyloid,
    metadata$Cing.Synuclein
  )
column_variables <-
  c(
    "Brain.wt",
    "Duration",
    "Age",
    "Cing.LB",
    "MF.LB",
    "Braak.NFT",
    "Thal.amyloid",
    "MF.SP",
    "MF.NFT",
    "MF.Amyloid",
    "Cing.Synuclein"
  )
APOE_comparison <-
  list(
    c("E2E3", "E2E4"),
    c("E2E3", "E3E3"),
    c("E2E3", "E3E4"),
    c("E2E3", "E4E4"),
    c("E2E4", "E3E3"),
    c("E2E4", "E3E4"),
    c("E2E4", "E4E4"),
    c("E3E3", "E3E4"),
    c("E3E3", "E4E4"),
    c("E3E4", "E4E4")
  )

violin_plot_fun <- function(i, j) {
  ggplot(metadata, aes(APOE, i, fill = APOE)) +
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.shape = NA) + 
    geom_jitter(shape=16, position=position_jitter(0.2)) +
    theme_bw() + 
    ggtitle(j) +
    xlab("APOE") +
    ylab(j) +
    stat_compare_means(comparisons = APOE_comparison) +
    scale_fill_manual(values=colorbindColors) +
    theme(legend.position = "none")
}
violin_plots <-
  Map(violin_plot_fun, i = metadata_continuous, j = column_variables)
violin_plots

plotnames = imap(violin_plots, ~paste0("../results/metadata/TYPE.APOE.", .y, ".pdf")) %>%
  flatten()
plotnames <- gsub("metadata.", "", plotnames)
walk2(plotnames, violin_plots, ~ggsave(filename = .x, plot = .y, 
                                    height = 4, width = 5.5))
remove(violin_plots) 
```

# Correlation heatmap
scale() function is used to ensure that variables measured in different scales (e.g. age of death vs RIN) are comparable.
same results with and without the use of scale
```{r correlation, message=FALSE, warning=FALSE, tidy=TRUE}
# scale continuous data 
metadata <-
  metadata[, c(
    "Brain.wt",
    "Age",
    "Braak.NFT",
    "Thal.amyloid",
    "Cing.LB"
  )] %>% scale()
info <- as.data.frame(metadata)
# correlation
cor_mat <- rcorr(as.matrix(info))
corrplot(
  cor_mat$r,
  method = "color",
  col = correlationColors(200),
  type = "upper",
  order = "hclust",
  p.mat = cor_mat$P,
  addCoef.col = "black",
  # Add coefficient of correlation
  tl.col = "black",
  tl.srt = 45,
  #Text label color and rotation
  diag = FALSE,
  col.lim = c(-1, 1)
)
path <- paste0("../results/metadata/correlation_continuous_variables_pathology_scores")
saveToPDF(paste0(path, ".pdf"), width = 5, height = 5)
```


